Investigating Neolithic pottery use from Guijiabao in the Yanyuan Basin in southwest China using organic residue analysis

### 2023-03-13 analysis/paper/paper.Rmd
minor edits

---
title: "Investigating Neolithic pottery use from Guijiabao in the Yanyuan Basin in southwest China using organic residue analysis"
author:
  - Li-Ying Wang:
      email: liyingw@asihp.net
      institute: [AcS]
      correspondence: true
  - Kuei-chen Lin:
      email: kueichenlin@asihp.net
      institute: [AcS]
      correspondence: false
  - Zhiqing Zhou:
      email: zhouzhiQ2000@163.com
      institute: [InsC]
      correspondence: false
institute:
  - AcS: Institute of History and Philology, Academia Sinica 
  - InsC: Chengdu Institute of Cultural Relics and Archaeology
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2:
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
bibliography: references.bib
csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
abstract: |
    The spread of new crops or species across areas during the Neolithic period is one of the important archaeological issues to understand the process of adoption of new practices. It is widely accepted that both millet and rice have been introduced into southwestern China at the latest in 4700 BP. However, to what extent the new crops were utilized by local cultures in an earlier time remains unclear. The excavation of Guijiabao in the Yanyuan basin in southwest China uncovered abundant materials with well-preserved stratigraphic records, providing a unique opportunity to understand the local adoption of agriculture. To further understand pottery use and food practices at Guijiabao, we investigated organic residues preserved in potsherds through lipid biomarker identification and compounds-specific stable isotope analysis. The results show the signals of terrestrial animals for most potsherds, with one showing a C3 plant source during the Neolithic period. Our case study indicates that hunting might have been the primary source of subsistence, supplemented by rice/millet farming. This study also demonstrates a perspective from organic residue analysis to explore pottery use and food practices during the Neolithic period. A more systematic sampling of potsherds across different types and temporal phases with larger samples is needed to provide a clear picture of ancient diets and subsistence.

@@

# Introduction
<!-- intro to the topic and research aim-->
Movements of people between geographically distinct areas usually accompany the exchange of goods, techniques, or ideas that could have cultural or economic impacts on those groups involved in the interaction [@Burmeister2000]. In southwestern China, the Hengduan Mountain ranges and valleys connecting the Qinghai-Tibetan Plateau and the Yunnan–Guizhou Plateau play an important role in the cultural exchange and spread of crops into this region and even further to mainland Southeast Asia [@Ma2021; @Lin2022; @Hao2022; @Tong1986]. Despite a long-standing debate between rice/millet dispersal waves from northwest China and the middle Yangtze valley [@Guedes2013; @DAlpoim2018; @Liu2022], it is widely accepted that mixed farming was practiced in southwestern Sichuan around 4,700 BP and to western Yunnan before 4600 BP [@Jiang2016; @YunnanMuseum1981; @Hao2022; @Zhang2010; @Dal2022; @Chen2015; @Weisskopf2015]. As a result, the Neolithic Hengduan Mountain areas are critical for understanding agricultural expansion processes. Recent archaeological evidence from the Yanyuan Basin, located in the middle Hengduan Mountains, reveals the presence of millet and rice at around the same time as mixed farming appeared in Sichuan [@Huan2022; @Hao2022]. However, to what extent the two crops were incorporated into the local diet is still not fully understood.

To explore whether the new crops or species became important food sources during the Neolithic period in the Yanyuan Basin, we investigated food residues from archaeological vessels at Guijiabao, a long-term occupation settlement with complete temporal sequences from the Neolithic period to the historical period in Yanyuan so far excavated [@Guijiabao2017a; @Guijiabao2020a]. The recent evidence at Guijiabao reveals that millet and rice were both present in the Neolithic period around 5000 BP, earlier than previously thought, and became part of the local diet [@Guijiabao2016; @Lin2022; @Huan2022]. Although some of these analyses suggest the presence of mixed millet/rice farming, we still lack direct evidence of culinary practices during this period. To understand the extent of new crops/species processing at Guijiabao, we ask the questions: What were the culinary practices reflected by pottery use? Could we identify the evidence of new crops or species incorporated into the local food practices during the Neolithic period? We answer those questions by examining potsherds using organic residue analysis, which allows us to directly explore food processed in ceramic vessels.

@@

Although the current findings at Guijiabao and Gaoshan support mixed agriculture, the proportions of millet or rice included in local culinary practices during the Neolithic period remain unclear. The macro-botanical and phytolith analyses for Guijiabao show that millet is the major crop, which is at odds with the predominance of rice at most sites in southwest China, such as Gaoshan and Baodun in Sichuan [@JiangGaoshan2017], and Henglanshan and Baiyangcun in Yunnan [@Hao2022]. The approach of residue analysis for ceramics reflecting food preparation and processing could be helpful to understand the adoption of new practices.

@@

## Methods
For organic residue analysis, we detected and identified lipid compounds by gas chromatography mass spectrometry (GC-MS) and quantified lipids by gas chromatography with flame-ionization detection (GC-FID). To distinguish the source of the common fatty acids, palmitic (C16:0) and stearic (C18:0), found in most animals and plants, we determined their stable carbon isotope characteristics by gas chromatography combustion isotope ratio mass spectrometry (GC-C-IRMS). The stable carbon isotope ratio for individual lipid compounds is a powerful indicator of the origins of fatty acids since different food categories, such as terrestrial animals, marine animals, and plants, have specific ranges of δ^13^C values [@Craig2004; @Craig2007; @Evershed1994].

For the sample processing, we cleaned the surface of each potsherd to prevent external contamination from environmental factors and handling during excavation or post-excavation analysis. Approximately 3.5 grams of powdered samples were extracted with a mixture of dichloromethane/methanol solution (9:1 v/v) using accelerated solvent extraction (ASE Dionex 350; at 100 °C). We conducted saponification by adding 10 mL of 1M KOH in methanol and 1 mL water to the extracts and heating them in an oven at 70 °C overnight. Saponified extracts were acidified with 6M hydrochloric acid (HCl), and the acid fraction was extracted using hexane ($3\times10$ mL). A 5% aliquot of each sample was derivatized using 20 uL N,O-bis(trimethylsilyl)trifluoroacetamide (BSTFA) with 1% v/v trimethylchlorosilane (TMCS), and heated at 70 °C for one hour. After drying under a stream of nitrogen, the derivatized sample was dissolved in toluene (20 uL) prior to GC-MS. Internal standards were added before and after the total lipid extraction.

The remaining 95% aliquot was fractioned by solid phase extraction (Aminopropyl, NH2, column with 0.5 g sorbent) with 8 mL dichloromethane:Isopropyl alcohol (3:1), 8 mL 4% acetic acid in ethyl ether, and 6 mL methanol, respectively. The second fraction was methylated to produce fatty acid methyl esters (FAMEs) with 1 mL dry hexane and 2 mL 10% (v/v) acetyl chloride in dry methanol, and heated at 70 °C overnight. The FAME derivatives were extracted with hexane ($3\times10$ mL) and then blown down with nitrogen streams. About 5% aliquot dissolved in toluene was analyzed by GC-FID for quantification, and the remaining sample was analyzed by GC-C-IRMS.

@@

```{r chromatogram-specific, cache= TRUE}
# plot the specific one
chroma_specific <-
  meth_all %>%
  filter(name %in% c("SYG-TN13-E22-2#2")) %>% 
  ggplot(aes(Path, File)) +
  geom_line(size = 0.3) +
  geom_text(data = highlight_SYG2_2, #ggrepel::geom_text_repel
            aes(label = FA),
            size = 3,
            nudge_y = 250000,
            show.legend = FALSE) +
  scale_y_continuous(labels = scales::comma_format(),
                     limits = c(0, 7500000),
                     breaks = seq(0, 7500000, 1000000)) +
  scale_x_continuous(limits = c(15, 50),
                     breaks = seq(15, 50, 5),
                     expand = c(0, 0.5)) + # don't log, many peaks and distort real counts
  labs(title = "SYG-TN13-E22-2#2", x = "retention time", y = "relative Intensity") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5),
        panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
        panel.background = element_blank(), axis.line = element_line(colour = "black"))

@@

To identify the sources of degraded animal fats, we conducted isotope analysis to obtain the stable carbon isotope compositions, δ^13^C values of C16:0 and C18:0, using GC-C-IRMS. The δ^13^C values of C16:0 and C18:0 have proven useful to distinguish between ruminant adipose, non-ruminant adipose, and ruminant dairy fats based on the different ways dietary carbon sources are utilized by different fat types [@Dunne2012; @Copley2003; @Evershed2008]. To reduce the bias caused by differential δ^13^C values across geographical regions, we used the Δ^13^C(δ^13^C18:0 - δ^13^C16:0) values to determine the sources of degraded fats [@Evershed2008; @Salque2013]. The carbon isotope values for C16:0 and C18:0 distribute between `r min_carbon_isotope` and `r max_carbon_isotope`, and their Δ13C values range from `r delta_min` ‰ to `r delta_max` ‰. The Δ13C values were plotted according to the global reference database for the identification of their origins [@Suryanarayan2021; @Dunne2012; @Evershed2008].

@@

# Discussion
<!--A brief summary of interesting results-->
The aim of this paper is to explore whether new crops, such as C4 plants (e.g., millets), C3 plants (e.g., rice), or new species (e.g., pigs/horses) became an essential part of the local diet in Yanyuan during the Neolithic period. Our results show that food choices indicated by lipid residues extracted from Guijiabao and Gaoshan potsherds were dominated by ruminant fats during the Neolithic period. Moreover, it is striking to identify non-ruminant fats in a Guijiabao potsherd that might originate from C3 plants. Although the lipid yields in most of our samples are generally low, interpretative amounts of yields from all of our samples support the assumption that those ceramics were used as cooking or storage pots. The low yield could be attributed to the base parts that we selected for analysis, which may absorb fewer residues than the rim of a vessel [@Evershed2008experimental; @Kwak2020; @Charters1993]. But we also found a base sample dating to the Neolithic period contained a rich composition of long-chain fatty acids, reflecting good preservation. We recognize limitations and challenges for interpreting lipid sources over time or differences across regions due to our small sample sizes and a lack of local modern references for animal fats and plant oils. Also, a detailed analysis of faunal remains from Guijiabao is still in progress, which constrains the comparison between different types of evidence. However, the general pattern of lipid profiles could still provide informative clues for a better understanding of culinary practices in Guijiabao.

<!--discussion about possible plant origins-->
While it has been indicated that millets (C4 plants), including foxtail and broomcorn millets, were the dominant cereals according to recent macro-botanical and phytolith studies at Guijiabao [@Guijiabao2016; @Hao2022; @Huan2022], we did not detect any lipid residues associated with C4 plants in our samples. Surprisingly, we identified non-ruminant fats from a Guijiabao potsherd that might be from terrestrial plants/C3 plants based on similar isotopic values reported by other studies in East Asia [@Shoda2018; @Kwak2020]. Based on the flotation results from Guijiabao, it is possible that the source of C3 plants is rice or native wild weeds, such as pennisetum and chenopodiaceae; however, the seeds of wild weeds only account for a small part of the plant [@Guijiabao2016]. Despite stone knives and tools were also found at the sites that might relate to farming practices, more studies are needed to confirm their usage for harvesting crops [@Zhou2019; @Hao2022; @Lin2022]. Another possible source for non-ruminant fats could be freshwater fish, which can be supported by the finding of fish bones and net sinkers at Guijiabao [@Lucquin2016; @Guijiabao2017b]. But we did not identify any lipid biomarkers related to aquatic resources, such as long-chain $\omega$-(o-alkylphenyl) alkanoic acids or phytanic acid, from our potsherds [@Evershed2008marine; @Hansel2004; @Cramp2014; @Kwak2017].

<!--discussion about possible ruminant fat origin-->
Ruminant fats identified in the Neolithic potsherds (ca. 4500 BP) from Guijiabao and Gaoshan could be derived from deer or horses, according to unearthed faunal remains from the sites and the local natural environments. Stone artifacts, including stone balls and arrowheads, were also found at the site, indicative of hunting activities [@Hao2022; @Guijiabao2020a]. However, whether horse was part of local food choices in southwest China around 4500 years ago is skeptical. First, the Pleistocene and Neolithic sites with wild horse bones are primarily concentrated in northern China, with only a sporadic presence in the south [@Yuan2006]. Second, domesticated horses first found in southwest China date to the Dian culture (4th century BC to 1st century AD) in eastern Yunnan [@Lei2009; @Chiou2004], which is 3000 years later than the Neolithic period we focus on. In addition, our sample (SYG-TN13-E22-2#1) shows an isotopic value consistent with the values of deer and sheep fats fed on a C3 diet based on global references [@Evershed2002]. Thus, deer is the most likely part of the local food choices. It is also possible that other ruminant animals, such as wild Asian water buffalo or yak, were food sources during this period [@Zhang2016].

We observed a mixture of ruminant fats and C3 plants based on a number of Guijiabao potsherds dating to the historical period (ca. 1000 BP). It appears that C3 plants could be processed more in the historical period compared to the Neolithic period. In addition to rice, the source of C3 plants could also include wheat since it was introduced during the historical period [@JiangGaoshan2017]. A larger sample size are needed to explore whether there is a shift in diet over time. The overall depleted δ^13^C values for potsherd samples associated with ruminant fats suggest that C3 plants might be an important food source for local animals. Despite direct C4 plant signals not being found, one Guijiabao potsherd demonstrates the source of C4 plant-fed ruminants during the Neolithic period. This indicates either C3 or C4 plants could be the food source for ruminants, but whether they were domesticated is unknown. The zooarchaeological analysis of faunal bones from Guijiabao is still in progress, which could provide more details about local animal species for comparison.

In general, during the Neolithic period, the food source detected in our pottery samples consisted mainly of ruminant fats, which are most likely from local wild animals such as deer. Considering the abundance of wild animal resources in the natural environment of the Hengduan mountain ranges, as suggested by the case of the Yingpanshan site [@He2020], a subsistence relying on hunting to a certain degree could be expected. Crop cultivation could be adopted as a supplement to local food sources while it was introduced in the Neolithic period, but it might only account for a small proportion. For the dietary pattern at Guijiabao previous study suggested a C4-based diet according to the isotopic analysis of human bones [@Lin2022], but we only identified a C3 plant signal from one pottery sample without any signal directly from C4 plants. This inconsistency could result from different sampling strategies or from the sample size issue, where our results only reflect a small portion of food content. Nevertheless, our results suggest that farming practices during the Neolithic period may not have been adopted as much as we previously thought. An organic residue approach may only provide a narrow perspective on local diets, but it could improve our current understanding of food practices through direct evidence that has been cooked in ceramics. A large-scale organic residue analysis of ceramics across different pottery forms in a wider region would offer insight into temporal and regional variations in the local diet during the Neolithic period.

# Conclusion
Our case study on organic residues preserved in pottery fragments from Guijiabao and Gaoshan in southwest China reveals the presence of lipids that could be associated with food types. The food processed in the ceramics included ruminant fats and terrestrial plants during the Neolithic period, when millet and rice spread into southwest China. The source of terrestrial plants could be C3 plants, implying the processing of rice or other wild weeds in the pot. However, the evidence of direct processing of either C3 or C4 plants is not in proportion to the large quantity of plant remains found at Guijiabao, especially millet grains. Instead, ruminant fats from local animals, such as wild deer, could have been a common food source during the early Neolithic. This suggests that hunting might have been an important part of local subsistence, supplemented by rice/millet farming when they were introduced. Our results from Guijiabao might represent a subsistence strategy for adapting to the natural environment of the Hengduan Mountains. We recognize the limitations of our interpretation of the local diet due to the small sample size. Nevertheless, our study could offer a perspective from organic residue analysis to explore food preferences and the extent of farming adoption when rice/millet agriculture first spread into the Yanyuan during the Neolithic period. A more systematic sampling of potsherds across different phases and pottery types may eliminate a clear picture of diet and subsistence during the Neolithic transition.

>>>>

---
title: "Investigating Neolithic pottery use from Guijiabao in the Yanyuan Basin in southwest China using organic residue analysis"
author:
  - Li-Ying Wang:
      email: liyingw@asihp.net
      institute: [AcS]
      correspondence: true
  - Kuei-chen Lin:
      email: kueichenlin@asihp.net
      institute: [AcS]
      correspondence: false
  - Zhiqing Zhou:
      email: zhouzhiQ2000@163.com
      institute: [InsC]
      correspondence: false
institute:
  - AcS: Institute of History and Philology, Academia Sinica 
  - InsC: Chengdu Institute of Cultural Relics and Archaeology
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2:
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
bibliography: references.bib
csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
abstract: |
    The spread of new crops or species across areas during the Neolithic period is one of the important archaeological issues to understand the process of adoption of new practices. It is widely accepted that both millet and rice have been introduced into southwest China at the latest in 4700 BP. However, to what extent the new crops were utilized by local cultures in an earlier time remains unclear. The excavation of Guijiabao in the Yanyuan basin in southwest China uncovered abundant materials with well-preserved stratigraphic records, providing a unique opportunity to understand the local adoption of agriculture. To further understand pottery use and food practices at Guijiabao, we investigated organic residues preserved in potsherds through lipid biomarker identification and compounds-specific stable isotope analysis. The results show the signals of terrestrial animals for most potsherds, with one showing a C3 plant source during the Neolithic period. Our case study indicates that hunting might have been the primary source of subsistence, supplemented by rice/millet farming. This study also demonstrates a perspective from organic residue analysis to explore pottery use and food practices during the Neolithic period. A more systematic sampling of potsherds across different types and temporal phases with larger samples is needed to provide a clear picture of ancient diets and subsistence.

@@

# Introduction
<!-- intro to the topic and research aim-->
Movements of people between geographically distinct areas usually accompany the exchange of goods, techniques, or ideas that could have cultural or economic impacts on those groups involved in the interaction [@Burmeister2000]. In southwestern China, the Hengduan Mountain ranges and valleys connecting the Qinghai-Tibetan Plateau and the Yunnan–Guizhou Plateau play an important role in the cultural exchange and spread of crops into this region and even further to mainland Southeast Asia [@Ma2021; @Lin2022; @Hao2022; @Tong1986]. Despite a long-standing debate between rice/millet dispersal waves from northwest China and the middle Yangtze valley [@Guedes2013; @DAlpoim2018; @Liu2022], it is widely accepted that mixed farming was practiced in southwest Sichuan around 4,700 BP and in western Yunnan before 4600 BP [@Jiang2016; @YunnanMuseum1981; @Hao2022; @Zhang2010; @Dal2022; @Chen2015; @Weisskopf2015]. The Neolithic Hengduan Mountain regions are therefore essential for understanding the processes of agricultural expansion. Recent archaeological evidence from the Yanyuan Basin in the middle Hengduan Mountains further reveals that millet and rice were present at roughly the same period as those grains found in Sichuan [@Huan2022; @Hao2022]. However, to what extent the two crops were incorporated into the local diet is still not fully understood.

To explore whether the new crops or species became important food sources during the Neolithic period in the Yanyuan Basin, we investigated food residues from archaeological vessels at Guijiabao, a long-term occupation settlement with complete temporal sequences from the Neolithic period to the historical period in Yanyuan so far excavated [@Guijiabao2017a; @Guijiabao2020a]. The recent evidence at Guijiabao reveals that millet and rice were both present in the Neolithic period around 5000 BP, earlier than previously thought [@Guijiabao2016; @Lin2022], and millet could have already become part of the local diet [@Lin2022]. Although some of these analyses suggest the presence of mixed millet/rice farming, we still lack direct evidence of culinary practices during this period. To understand the extent of new crops/species processing at Guijiabao, we ask the questions: What were the culinary practices reflected by pottery use? Could we identify the evidence of new crops or species incorporated into the local food practices during the Neolithic period? We answer those questions by examining potsherds using organic residue analysis, which allows us to directly explore food processed in ceramic vessels.

@@

Although the recent findings at Guijiabao and Gaoshan support mixed agriculture, it is still unclear how much millet or rice was included in local culinary practices during the Neolithic period. The macro-botanical and phytolith analyses for Guijiabao show that millet is the major crop, similar to the findings at the Yingpanshan site in Sichuan [@Zhao2011]. However, plant remains from many sites in southwest China reveal a predominance of rice, such as Gaoshan and Baodun in Sichuan [@JiangGaoshan2017] and Henglanshan and Baiyangcun in Yunnan [@Hao2022]. This indicates there were variations in agricultural practices in southwest China. To better understand the extent of the adoption of new crops, organic residue analysis for ceramics is a useful approach since we can directly study food residues related to culinary practices, including food preparation, processing, and cooking. This provides another way to examine the expansion of agriculture in the Yanyuan area.

@@

## Methods
We detected and identified lipid compounds by gas chromatography mass spectrometry (GC-MS) and quantified lipids by gas chromatography with flame-ionization detection (GC-FID). To distinguish the source of the common fatty acids, palmitic (C16:0) and stearic (C18:0), found in most animals and plants, we determined their stable carbon isotope characteristics by gas chromatography combustion isotope ratio mass spectrometry (GC-C-IRMS). The stable carbon isotope ratio for individual lipid compounds is a powerful indicator of the origins of fatty acids since different food categories, such as terrestrial animals, marine animals, and plants, have specific ranges of δ^13^C values [@Craig2004; @Craig2007; @Evershed1994].

For the sample processing, the surface of each potsherd was removed to prevent external contamination from environmental factors and handling during excavation or post-excavation analysis. Approximately 3.5 grams of powdered samples were extracted with a mixture of dichloromethane/methanol solution (9:1 v/v) using accelerated solvent extraction (ASE Dionex 350; at 100 °C). We conducted saponification by adding 10 mL of 1M KOH in methanol and 1 mL water to the extracts and heating them in an oven at 70 °C overnight. Saponified extracts were acidified with 6M hydrochloric acid (HCl), and the acid fraction was extracted using hexane ($3\times10$ mL). A 5% aliquot of each sample was derivatized using 20 uL N,O-bis(trimethylsilyl)trifluoroacetamide (BSTFA) with 1% v/v trimethylchlorosilane (TMCS), and heated at 70 °C for one hour. After drying under a stream of nitrogen, the derivatized sample was dissolved in toluene (20 uL) prior to GC-MS.

The remaining 95% aliquot was fractioned by solid phase extraction (Aminopropyl, NH2, column with 0.5 g sorbent) with 8 mL dichloromethane:Isopropyl alcohol (3:1), 8 mL 4% acetic acid in ethyl ether, and 6 mL methanol, respectively. The second fraction was methylated to produce fatty acid methyl esters (FAMEs) with 1 mL dry hexane and 2 mL 10% (v/v) acetyl chloride in dry methanol, and heated at 70 °C overnight. The FAME derivatives were extracted with hexane ($3\times10$ mL) and then blown down with nitrogen streams. About 5% aliquot dissolved in toluene was analyzed by GC-FID for quantification, and the remaining sample was analyzed by GC-C-IRMS. With each batch of samples, we prepared analytical blanks to check for potential contamination sources that might from solvent and reagent. Internal standards were added before and after the lipid extraction for quantification.

@@

```{r chromatogram-specific, cache= TRUE}
# plot the specific one
chroma_specific <-
  meth_all %>%
  filter(name %in% c("SYG-TN13-E22-2#2")) %>% 
  ggplot(aes(Path, File)) +
  geom_line(size = 0.3) +
  ggrepel::geom_text_repel(data = highlight_SYG2_2, #ggrepel::geom_text_repel
            aes(label = FA),
            size = 3.5,
            nudge_y = 250000,
            show.legend = FALSE) +
  scale_y_continuous(labels = scales::comma_format(),
                     limits = c(0, 7500000),
                     breaks = seq(0, 7500000, 1000000)) +
  scale_x_continuous(limits = c(15, 50),
                     breaks = seq(15, 50, 5),
                     expand = c(0, 0.5)) + # don't log, many peaks and distort real counts
  labs(title = "SYG-TN13-E22-2#2", x = "retention time", y = "relative Intensity") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5),
        panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
        panel.background = element_blank(), axis.line = element_line(colour = "black"))

@@

To identify the sources of degraded animal fats, we conducted isotope analysis to obtain the stable carbon isotope compositions, δ^13^C values of C16:0 and C18:0, using GC-C-IRMS. The δ^13^C values of C16:0 and C18:0 have proven useful to distinguish between ruminant adipose, non-ruminant adipose, and ruminant dairy fats based on the different ways dietary carbon sources are utilized by different fat types [@Dunne2012; @Copley2003; @Evershed2008]. To reduce the bias caused by differential δ^13^C values across geographical regions, we used the Δ^13^C (δ^13^C18:0 - δ^13^C16:0) values to determine the sources of degraded fats [@Evershed2008; @Salque2013]. The carbon isotope values for C16:0 and C18:0 distribute between `r min_carbon_isotope` and `r max_carbon_isotope`, and their Δ13C values range from `r delta_min` ‰ to `r delta_max` ‰. The Δ13C values were plotted according to the global reference database for the identification of their origins [@Suryanarayan2021; @Dunne2012; @Evershed2008].

@@

# Discussion
<!--A brief summary of interesting results-->
The aim of this paper is to explore whether new crops, such as C4 plants (e.g., millets), C3 plants (e.g., rice), or new species (e.g., pigs/horses) became an essential part of the local diet in Yanyuan during the Neolithic period. Our results from lipid residues in Guijiabao and Gaoshan potsherds show that food categories were dominated by ruminant fats during the Neolithic period. Moreover, it is striking to identify non-ruminant fats in a Guijiabao potsherd that might originate from C3 plants (see below). Although the lipid yields in most of our samples are generally low, interpretative amounts of yields from all of our samples support the assumption that those ceramics were used as cooking or storage pots. The low yield could be attributed to the bottom parts of the vessels for analysis, which may absorb fewer residues than the rim of a vessel [@Evershed2008experimental; @Kwak2020; @Charters1993]. But we also found that one bottom sample dated to the Neolithic period contains a richer composition of long-chain fatty acids compared with those in other samples, reflecting better preservation or a much higher lipid yield. We recognize limitations and challenges for interpreting lipid sources over time or differences across regions due to our small sample sizes and a lack of local modern references for animal fats and plant oils. Besides, a detailed analysis of faunal remains from Guijiabao is still in progress, which may constrain a comprehensive comparison between different types of evidence. Nevertheless, the general pattern of lipid profiles could still provide informative clues for a better understanding of culinary practices in Guijiabao.

<!--discussion about possible plant origins-->
While it has been indicated that millets (C4 plants), including foxtail and broomcorn millets, were the dominant cereals according to recent macro-botanical and phytolith studies at Guijiabao [@Guijiabao2016; @Hao2022; @Huan2022], we did not detect any lipid residues associated with C4 plants in our samples. Surprisingly, we identified non-ruminant fats from a Guijiabao potsherd that might be from terrestrial plants/C3 plants based on similar isotopic values reported by other studies in East Asia [@Shoda2018; @Kwak2020]. According to the flotation results from Guijiabao, it is possible that the source of C3 plants is rice or native wild weeds, such as pennisetum and chenopodiaceae; however, the seeds of wild weeds only account for a small part of the floral assemblage derived from flotation [@Guijiabao2016]. Despite stone knives and tools were also found at the sites that might relate to farming practices, more studies are needed to confirm their usage for harvesting crops [@Zhou2019; @Hao2022; @Lin2022]. Another possible source for non-ruminant fats could be freshwater fish, which can be supported by the finding of fish bones and net sinkers at Guijiabao [@Guijiabao2017b]. But we did not identify any lipid biomarkers related to aquatic resources, such as long-chain $\omega$-(o-alkylphenyl) alkanoic acids or phytanic acid, from our potsherds [@Evershed2008marine; @Hansel2004; @Cramp2014; @Lucquin2016].

<!--discussion about possible ruminant fat origin-->
The ruminant fats identified in the Neolithic potsherds (ca. 4500 BP) from Guijiabao and Gaoshan could be derived from deer, sheep, or horses fed on a C3 diet based on their isotopic values that are consistent with global references [@Evershed2002; @Mileto2017]. This also corresponds with unearthed faunal remains from the sites and the local natural environments. Stone artifacts found at the site, including stone balls and arrowheads, further indicates hunting activities [@Hao2022; @Guijiabao2020a]. However, whether horse was part of local food choices in southwest China around 4500 years ago is skeptical. First, the Pleistocene and Neolithic sites with wild horse bones are primarily concentrated in northern China, with only a sporadic presence in the south [@Yuan2006]. Second, domesticated horses first found in southwest China date to the Dian culture (4th century BC to 1st century AD) in eastern Yunnan [@Lei2009; @Chiou2004], which is 3000 years later than the Neolithic period we focus on. Thus, supported by faunal evidence, deer is the most likely part of the local food choices. It is also possible that local wild bovids, such as yaks or water buffalo, were food sources during this period, as these species have been found in southwest China [@Zhang2016; @Chen2020].

We observed a mixture of ruminant fats and C3 plants based on a number of Guijiabao potsherds that date to the historical period (ca. 1000 BP). It appears that C3 plants could be processed more in the historical period than they were in the Neolithic period. In addition to rice, the source of C3 plants could also include wheat since it was introduced during the historical period [@JiangGaoshan2017]. A larger sample size is needed to explore whether there is a shift in diet over time. The overall depleted δ^13^C values for potsherd samples associated with ruminant fats suggest that C3 plants might be an important food source for local animals. Despite direct C4 plant signals not being found, one Guijiabao potsherd demonstrates the source of C4 plant-fed ruminants during the Neolithic period. This indicates either C3 or C4 plants could be the food source for ruminants, but whether those animals were domesticated is unknown. The zooarchaeological analysis of faunal bones from Guijiabao is still in progress, which could provide more details about local animal species for comparison.

In general, ruminant fats from local wild animals, such as deer, were the predominant food source during the Neolithic period based on our results. A subsistence that relied on hunting to a certain degree could be anticipated given the abundance of wild animals in the natural settings of the Hengduan mountain ranges, as indicated by the faunal remains from the Yingpanshan site [@He2020]. Crop cultivation could be adopted as a supplement to local food sources while it was introduced in the Neolithic period, but it might only account for a small proportion. For the dietary pattern at Guijiabao, previous study suggested a C4-based diet according to the isotopic analysis of human bones [@Lin2022], but we only identified a C3 plant signal from one pottery sample without any signal directly from C4 plants. This inconsistency could result from different sampling strategies or from the sample size issue, where our results only reflect a small portion of food content. Nevertheless, our results suggest that farming practices during the Neolithic period may not have been adopted as much as we previously thought. An organic residue approach may only provide a narrow perspective on local diets, but it could improve our current understanding of food practices through direct evidence that has been cooked in ceramics. A large-scale organic residue analysis of ceramics across different pottery forms in a wider region would offer insight into temporal and regional variations in the local diet during the Neolithic period.

# Conclusion
Our case study on organic residues in pottery fragments from Guijiabao and Gaoshan in southwest China reveals the presence of lipids that was associated with food categories. The food processed in the ceramics included ruminant fats and terrestrial plants during the Neolithic period, when millet and rice spread into southwest China. The source of terrestrial plants could be C3 plants, implying the processing of rice or other wild weeds in the pot. However, the evidence of direct processing of either C3 or C4 plants is not in proportion to the large quantity of plant remains found at Guijiabao, especially millet grains. Instead, ruminant fats from local animals, such as wild deer, could have been a common food source during the Neolithic. This suggests that hunting might have been an important part of local subsistence, supplemented by rice/millet farming when they were introduced. Our results from Guijiabao could represent a subsistence strategy for adapting to the natural environment of the Hengduan Mountains. We recognize the limitations of our interpretation of the local diet due to the small sample size. Nevertheless, our study could offer a perspective from organic residue analysis to explore food choices and the extent of farming adoption when rice/millet agriculture first spread into the Yanyuan during the Neolithic period. A more systematic sampling of potsherds across different phases and pottery types may eliminate a clear picture of diet and subsistence during the Neolithic transition.
